---
title: "Getting started with shinytone"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with shinytone}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(shinytone)
library(dplyr)
library(ggplot2)
```

## What is shinytone?

**shinytone** is a research hub for *citation tone analysis* in tone languages.
It provides:

* An interactive Shiny app that walks you through the full workflow — pitch
  extraction, by-speaker f0 normalisation, outlier detection, growth-curve
  and generalised additive mixed models, and Chao tone numeral summarisation.
* A small set of pure R functions that implement the analytical core, so you
  can run the same analyses scripted from RMarkdown or another package.

This vignette walks through the package's scripted side using a small toy
dataset. If you prefer a graphical workflow, install the package and run
`shinytone::run_app()`, or visit the hosted app at
<https://chenzixu.shinyapps.io/shinytone/>.

## A toy dataset

The package functions expect long-format data — one row per f0 sample. For
this vignette we build a tiny synthetic corpus: 2 speakers × 3 tones × 4
tokens per tone × 20 frames per token.

```{r}
set.seed(1)

make_toy_corpus <- function(n_frames = 20) {
  spec <- expand.grid(
    speaker = c("S01", "S02"),
    tone    = c("T1", "T2", "T3"),
    rep     = 1:4,
    stringsAsFactors = FALSE
  )
  spec$token <- paste(spec$speaker, spec$tone, spec$rep, sep = "_")
  spec$item  <- paste0("w", spec$rep)

  do.call(rbind, lapply(seq_len(nrow(spec)), function(i) {
    r <- spec[i, ]
    t <- seq(0, 1, length.out = n_frames)
    # Speaker baseline: female speaker S01 a bit higher than S02
    base <- if (r$speaker == "S01") 220 else 130
    # Tone shape: T1 level high, T2 rising, T3 falling
    shape <- switch(r$tone,
      T1 = rep(0,  n_frames),
      T2 = 20 * t,
      T3 = -25 * t
    )
    data.frame(
      speaker = r$speaker,
      tone    = r$tone,
      item    = r$item,
      token   = r$token,
      time    = t,
      f0      = base + shape + rnorm(n_frames, sd = 3),
      stringsAsFactors = FALSE
    )
  }))
}

toy <- make_toy_corpus()
head(toy)
```

Two speakers means the absolute Hz values differ — that's the situation
`normalise_f0()` is designed for.

## Normalise f0 by speaker

`normalise_f0()` adds a `speaker_mean` column plus either `f0_st` (semitones,
default) or `f0_zscore`, computed per speaker:

```{r}
normed <- normalise_f0(toy,
                       f0          = "f0",
                       speaker     = "speaker",
                       tone        = "tone",
                       method      = "semitone",
                       mean_method = "weighted")
head(normed)
```

Now `f0_st` is on a comparable scale across speakers. A quick visual:

```{r}
ggplot(normed, aes(time, f0_st, colour = tone, group = token)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ speaker) +
  labs(y = "f0 (semitones, relative to speaker mean)")
```

## Inspect for outliers and pitch-tracking artefacts

`inspect_f0()` flags two kinds of issues: tokens whose per-token max or min
sits more than `z_threshold` SDs from the speaker's mean, and individual
samples where the rate of f0 change exceeds physiological plausibility
(Sundberg 1973, Steffman & Cole 2022).

```{r}
inspected <- inspect_f0(toy,
                        f0      = "f0",
                        token   = "token",
                        time    = "time",
                        speaker = "speaker",
                        tone    = "tone")
inspected |>
  distinct(token, .keep_all = TRUE) |>
  count(flagged_token)
```

For the toy data, nothing is flagged (it's too clean). On real data,
`flagged_token` and the human-readable `flag_notes` column tell you which
tokens to investigate.

## Fit token-level polynomials

`fit_polynomial()` returns one row per token with Legendre polynomial
coefficients (`c0`, `c1`, `c2`, ...). Useful as inputs to downstream
classifiers or regression models.

```{r}
poly_coefs <- fit_polynomial(normed,
                             f0      = "f0_st",
                             token   = "token",
                             time    = "time",
                             speaker = "speaker",
                             tone    = "tone",
                             degree  = 2)
head(poly_coefs)
```

`c0` ≈ token-level mean f0 (in semitones). `c1` ≈ linear slope (positive
for rising, negative for falling). `c2` ≈ curvature.

## Fit a Growth Curve Analysis (GCA)

`fit_gca()` is a thin wrapper around `lme4::lmer()` with sensible defaults
for tone-shape modelling — orthogonal polynomials on per-token normalised
time, plus conventional random effects on speaker and item.

```{r, eval = FALSE}
gca <- fit_gca(normed,
               f0      = "f0_st",
               time    = "time",
               token   = "token",
               tone    = "tone",
               speaker = "speaker",
               item    = "item",
               degree  = 2,
               random_slope_speaker = FALSE,
               random_slope_item    = FALSE)

# Population-level per-tone curves
preds <- predict_gca(gca, n = 60)
ggplot(preds, aes(time, f0_predicted, colour = tone)) +
  geom_line(linewidth = 1)
```

For non-standard model structures (custom contrasts, alternative random
effects), call `lme4::lmer()` directly — `fit_gca()` is just a convenience
wrapper for the common case.

## Convert to Chao tone numerals

`contour_to_chao()` reduces per-tone mean (or model-predicted) contours to
Chao numerals — 5-level digit strings like `55`, `35`, or `214`.

```{r}
# Pre-aggregate the toy data into a per-tone mean contour
mean_contour <- compute_mean_contour(toy,
                                     token = "token", f0 = "f0",
                                     time  = "time",  tone = "tone")

chao <- contour_to_chao(mean_contour, raw_data = toy,
                        raw_token = "token", raw_f0 = "f0",
                        raw_tone  = "tone")
chao[, c("tone", "refline", "interval", "robust", "shape")]
```

Three conversion methods are computed in one pass:

* **`refline`** — reference-line FOR, rounded on the `[1, 5]` scale
* **`interval`** — interval-based FOR, ceiling on the `(0, 5]` scale
* **`robust`** — robust FOR using μ ± σ of the highest- and lowest-tone
  extremes (requires `raw_data`)

The `shape` column summarises the contour ("rising", "falling", "dipping",
etc.) and is derived from the reference-line numeral via
`classify_contour()`.

## Where to go next

* Browse the [function reference](../reference/index.html) for the full API.
* Run `shinytone::run_app()` locally for the full graphical workflow,
  including audio handling and a Praat script generator.
* Cite the package with `citation("shinytone")` in any work that uses it.

## References

* Steffman, J., & Cole, J. (2022). Pitch tracking artefacts and the
  detection of voicing errors in spontaneous speech.
* Sundberg, J. (1973). Data on maximum speed of pitch changes.
* Xu, C. (2025). Tone production in Changsha. *Phonetica*.
