---
title: "Getting started with shinytone"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with shinytone}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  message   = FALSE,
  warning   = FALSE
)
suppressPackageStartupMessages({
  library(shinytone)
  library(dplyr)
  library(ggplot2)
})
```

## What is shinytone?

**shinytone** is a research hub for *citation tone analysis* in tone
languages. It bundles:

* An interactive Shiny app that walks you through the full workflow —
  pitch extraction, by-speaker f0 normalisation, outlier detection,
  growth-curve and generalised additive mixed models, and Chao tone
  numeral summarisation.
* A small set of pure R functions that implement the analytical core, so
  you can run the same analyses scripted from RMarkdown or another package.

This vignette walks through the scripted side using the package's bundled
sample dataset. If you prefer a graphical workflow, run
`shinytone::run_app()` after installing, or visit the hosted app at
<https://chenzixu.shinyapps.io/shinytone/>.

## Setup

The examples below assume these three packages are loaded:

```{r setup_visible, eval = FALSE}
library(shinytone)
library(dplyr)
library(ggplot2)
```

## The bundled `sample_f0` dataset

shinytone ships with a real citation-tone corpus available immediately
after `library(shinytone)`. It contains 38,808 f0 samples from 1,848
tokens across 13 speakers.

```{r}
data(sample_f0)
glimpse(sample_f0)
```

The columns we'll use:

* `token` — unique identifier for each recorded syllable (groups rows of
  the same contour together)
* `time` — time in seconds within each token
* `f0_Hz` — fundamental frequency
* `speaker` — speaker ID
* `tone` — Chao tone category (1-5)
* `char` — the Chinese character of the syllable

See `?sample_f0` for the full schema and source citation.

## Normalise f0 by speaker

Raw Hz isn't comparable across speakers (men and women have very different
baseline f0). `normalise_f0()` adds a `speaker_mean` column plus either
`f0_st` (semitones, default) or `f0_zscore`, computed per speaker:

```{r}
normed <- normalise_f0(sample_f0,
                       f0          = "f0_Hz",
                       speaker     = "speaker",
                       tone        = "tone",
                       method      = "semitone",
                       mean_method = "weighted")
head(normed[, c("speaker", "tone", "f0_Hz", "speaker_mean", "f0_st")])
```

Now `f0_st` puts every speaker on a comparable scale. To plot the mean
contour per tone, use [compute_mean_contour()], which normalises time
to `[0, 1]` *within each token* before averaging. The package function
returns its result in the same `time / f0_predicted / tone` schema used
by [predict_gca()] and [predict_gamm()]:

```{r}
mean_contour <- compute_mean_contour(normed,
                                     token = "token", f0 = "f0_st",
                                     time  = "time",  tone = "tone")

ggplot(mean_contour, aes(time, f0_predicted, colour = factor(tone))) +
  geom_line(linewidth = 1) +
  labs(x = "Normalised time", y = "f0 (semitones)", colour = "Tone")
```

## Inspect for outliers and pitch-tracking artefacts

`inspect_f0()` runs three complementary checks: tokens whose per-token
maximum sits more than `z_threshold` SDs above, or whose minimum more than
`z_threshold` SDs below, the speaker's other tokens (extreme-value), tokens
whose median is unusual for their speaker and tone (token-level), and
individual samples where the rate of f0 change exceeds physiological
plausibility (sample-level; Sundberg 1973, Steffman & Cole 2022).

The token-level check is tone-relative, so it runs only when a `tone` column
is available. If your tone categories are not yet known (for example before
the clustering / tone-discovery step, `cluster_f0()`), call `inspect_f0()`
with `tone = NULL`: the speaker-level extreme-value and sample-level jump
checks still run, and only the token-level check is skipped, with the `tone`
column omitted from the output.

```{r}
inspected <- inspect_f0(sample_f0,
                        f0      = "f0_Hz",
                        token   = "token",
                        time    = "time",
                        speaker = "speaker",
                        tone    = "tone")

# How many tokens were flagged overall?
inspected |>
  distinct(token, .keep_all = TRUE) |>
  count(flagged_token)
```

The `flag_notes` column gives a human-readable reason for each flagged
sample (e.g., `"max too high; jump (rise)"`). On the live app, the
Inspect tab lets you click through flagged tokens one by one.

## Fit token-level polynomials

`fit_polynomial()` returns one row per token with Legendre polynomial
coefficients (`c0`, `c1`, `c2`, ...). Useful as features for downstream
classifiers, or as a compact summary of contour shape.

```{r}
poly_coefs <- fit_polynomial(normed,
                             f0      = "f0_st",
                             token   = "token",
                             time    = "time",
                             speaker = "speaker",
                             tone    = "tone",
                             degree  = 2)
head(poly_coefs)
```

Interpretation:

* `c0` ≈ token-level mean f0 (in semitones, relative to speaker mean)
* `c1` ≈ linear slope across the contour (positive = rising)
* `c2` ≈ curvature (positive = U-shaped, negative = inverted-U)

## Fit a Growth Curve Analysis (GCA)

`fit_gca()` is a convenience wrapper around `lme4::lmer()` with sensible
defaults for tone-shape modelling — orthogonal polynomials on per-token
normalised time, plus conventional random effects on speaker and item.
For non-standard model structures, call `lme4::lmer()` directly.

```{r, eval = FALSE}
# (Skipped on package build for speed; uncomment to run locally.)
gca <- fit_gca(normed,
               f0      = "f0_st",
               time    = "time",
               token   = "token",
               tone    = "tone",
               speaker = "speaker",
               item    = "char",                 # use Chinese char as item
               degree  = 2,
               random_slope_speaker = FALSE,
               random_slope_item    = FALSE)

# Population-level per-tone curves
preds <- predict_gca(gca, n = 100)
ggplot(preds, aes(time, f0_predicted, colour = tone)) +
  geom_line(linewidth = 1)
```

## Fit a GAMM and check its diagnostics

`fit_gamm()` wraps `mgcv::bam()` with the smooth structure recommended
for tone contours (Sóskuthy 2017): a reference smooth plus per-tone
difference smooths, by-speaker random smooths, and an optional AR1
correction for the strong within-token autocorrelation of
densely-sampled f0 (`use_ar1 = TRUE`; the app's GAMM tab enables it by
default). The AR1 parameter rho is estimated from the lag-1
autocorrelation of the residuals *within* tokens, so token boundaries
do not bias it. `diagnose_gamm()` then answers the model-checking
questions in one call: is the basis dimension `k` large enough, are the
residuals well-behaved, and did the AR1 correction actually remove the
autocorrelation.

```{r, eval = FALSE}
# (Skipped on package build for speed; uncomment to run locally.)
gamm <- fit_gamm(normed,
                 f0      = "f0_st",
                 time    = "time",
                 token   = "token",
                 tone    = "tone",
                 speaker = "speaker",
                 item    = "char",
                 use_ar1 = TRUE)

# Population-level per-tone curves, same schema as predict_gca()
preds <- predict_gamm(gamm, n = 100)
ggplot(preds, aes(time, f0_predicted, colour = tone)) +
  geom_line(linewidth = 1)

# Model checking: basis-dimension (k) check, residual panels, per-token
# residual ACF (AR1-whitened when use_ar1 = TRUE), and concurvity
diag <- diagnose_gamm(gamm)
diag$k_check
head(diag$acf)
```

## Convert to Chao tone numerals

`contour_to_chao()` reduces per-tone mean (or model-predicted) contours
into Chao numerals — 5-level digit strings like `55`, `35`, or `214`.

```{r}
# Pre-aggregate into a per-tone mean contour over all speakers
mean_contour <- compute_mean_contour(sample_f0,
                                     token = "token", f0 = "f0_Hz",
                                     time  = "time",  tone = "tone")

chao <- contour_to_chao(
  mean_contour,
  raw_data  = sample_f0,
  raw_token = "token", raw_f0 = "f0_Hz", raw_tone = "tone"
)
chao[, c("tone", "refline", "interval", "robust", "shape")]
```

Three conversion methods in one pass:

* **`refline`** — reference-line FOR, rounded on the `[1, 5]` scale
* **`interval`** — interval-based FOR, ceiling on the `(0, 5]` scale
* **`robust`** — robust FOR using μ ± σ of the highest- and lowest-tone
  extremes (requires `raw_data`)

The `shape` column summarises the contour ("rising", "falling",
"dipping", etc.) and is derived from the reference-line numeral via
`classify_contour()`.

## Where to go next

* Browse the [function reference](https://chenchenzi.github.io/citationtone_hub/reference/) for the full API.
* Run `shinytone::run_app()` locally for the full graphical workflow,
  including audio processing and a Praat script generator.
* Cite the package with `citation("shinytone")` in any work that uses it.

## References

* Sóskuthy, M. (2017). Generalised additive mixed models for dynamic
  analysis in linguistics: A practical introduction. arXiv:1703.05339.
* Steffman, J., & Cole, J. (2022). Pitch tracking artefacts and the
  detection of voicing errors in spontaneous speech.
* Sundberg, J. (1973). The acoustics of the singing voice.
  *Scientific American*, 229(3), 82–91.
* Xu, C. (2025). Plastic Mandarin tones: regional identity in prosody.
  *Phonetica*, 82(5), 331–362.
  <https://doi.org/10.1515/phon-2025-0001>

* Xu, C., & Zhang, C. (2024). A cross-linguistic review of citation
  tone production studies: Methodology and recommendations. *The
  Journal of the Acoustical Society of America*, 156(4), 2538–2565.
  <https://doi.org/10.1121/10.0032356>
